Redirecting Resolution Proofs

### 2011-05-10 bib/prooftransform/redirecting-proofs-2/paper.tex
ispell

Our interest in readable ATP proofs has a different origin. The tool
Sledgehammer \cite{meng-paulson-2008,paulson-susanto-2007} integrates
ATPs with the interactive theorem prover Isabelle/HOL
\cite{nipkow-et-al-2002}. Given an Isabelle/HOL conjecture,
Sledgehammer heuristically selects a few hundred relevant lemmas from
Isabelle's libraries, translates them along with the conjecture, and
sends the resulting problem to E~\cite{schulz-2004},
SPASS~\cite{weidenbach-2001}, and
Vampire~\cite{riazanov-voronkov-2002}.
%
Proofs are reconstructed in Isabelle either using a single invocation
of the built-in resolution prover Metis \cite{hurd-2003} or as
structured Isar \cite{wenzel-2007} proofs following the ATP proofs
\cite{paulson-susanto-2007}. This latter option is useful for larger
proofs, which Metis fails to re-find within a reasonable time. But
most users find the proofs hideous and are little enclined to
insert them in their formalizations.

@@

Here we were lucky because we could join both branches the node 2 to avoid all
duplication. If we want to stick to a strick no-duplication policy, in general
we sometimes need to join on a disjunction $\varphi_1 \lor \cdots \lor
\varphi_n$, as the next example will illustrate.

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
 & [&5] & \\
{[2]} & $5 \have {}$ & $2 \lor 4$ & [6]
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more aesthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.

>>>>

Our interest in readable ATP proofs has a different origin. The tool
Sledgehammer \cite{meng-paulson-2008,paulson-susanto-2007} integrates
ATPs with the interactive theorem prover Isabelle/HOL
\cite{nipkow-et-al-2002}. Given an Isabelle/HOL conjecture,
Sledgehammer heuristically selects a few hundred relevant lemmas from
Isabelle's libraries, translates them along with the conjecture, and
sends the resulting problem to E~\cite{schulz-2004},
SPASS~\cite{weidenbach-2001}, and
Vampire~\cite{riazanov-voronkov-2002}.
%
Proofs are reconstructed in Isabelle either using a single invocation
of the built-in resolution prover Metis \cite{hurd-2003} or as
structured Isar \cite{wenzel-2007} proofs following the ATP proofs
\cite{paulson-susanto-2007}. This latter option is useful for larger
proofs, which Metis fails to re-find within a reasonable time. But
most users find the proofs hideous and are little inclined to
insert them in their formalizations.

@@

Here we were lucky because we could join both branches the node 2 to avoid all
duplication. If we want to stick to a strict no-duplication policy, in general
we sometimes need to join on a disjunction $\varphi_1 \lor \cdots \lor
\varphi_n$, as the next example will illustrate.

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
 & [&5] & \\
{[2]} & $5 \have {}$ & $2 \lor 4$ & [6]
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more esthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.
